BarnebyLives: an R package to create herbarium specimen labels and clean spreadsheets

### 2025-09-20 LabelStyles/SoS-skeleton-Spaced.Rmd
working on vignette example
reviewer comments on linux mac

```{r load data}
data <- read.csv('SoS-ExampleCollection.csv') |>
  dplyr::mutate(
    Collection_number = as.numeric(Collection_number), 
    Coordinate_Uncertainty = '+/- 5m') |>
  dplyr::filter(Collection_number == params$Collection_number)
```

\begin{center}
\textbf{\LARGE Seeds of Success}
\end{center}
\begin{flushright}
\uppercase{`r data$Family`}
\end{flushright}
\begingroup
\large
`r writer(paste(data$Genus, data$Epithet), italics = TRUE)` `r writer(data$Binomial_authority)` `r writer(data$Infrarank)` `r  writer(data$Infraspecies, italics = TRUE)` `r writer(data$Infraspecific_authority)`  
\normalsize
`r writer_fide(data)`
\endgroup

\vspace*{\fill}
\begingroup
`r data$Gen`. `r writer(data$Site)` `r data$latitude_dd` `r data$longitude_dd` (`r data$Datum` `r writer(data$Coordinate_Uncertainty)`).

@@

\vspace*{\fill}
\begingroup
`r collection_writer(data)`.
\endgroup

>>>>

```{r load data}
data <- read.csv('SoS-ExampleCollection.csv') |>
  dplyr::mutate(
    Collection_number = as.numeric(Collection_number), 
    Coordinate_uncertainty = '+/- 5m') |>
  dplyr::filter(Collection_number == params$Collection_number)
```

\begin{center}
\textbf{\LARGE Seeds of Success}
\end{center}
\begin{flushright}
\uppercase{`r data$Family`}
\end{flushright}
\begingroup
\large
`r writer(paste(data$Genus, data$Epithet), italics = TRUE)` `r writer(data$Binomial_authority)` `r writer(data$Infrarank)` `r  writer(data$Infraspecies, italics = TRUE)` `r writer(data$Infraspecific_authority)`  
\normalsize
\endgroup

\vspace*{\fill}
\begingroup
`r data$Gen`. `r writer(data$Site)` `r data$latitude_dd` `r data$longitude_dd` (`r data$Datum` `r writer(data$Coordinate_uncertainty)`).

@@

\vspace*{\fill}
\begingroup
`r collection_writer(data)`. `r writer_fide(data)`
\endgroup

### 2025-09-20 LabelStyles/SoS-skeleton-familyLeft.Rmd
reviewer comments on linux mac

```{r load data}
data <- read.csv('SoS-ExampleCollection.csv') |>
  dplyr::mutate(
    Collection_number = as.numeric(Collection_number), 
    Coordinate_Uncertainty = '+/- 5m') |>
  dplyr::filter(Collection_number == params$Collection_number)
```

@@

\vspace*{\fill}
\begingroup
`r data$Gen`. `r writer(data$Site)` `r data$latitude_dd` `r data$longitude_dd` (`r data$Datum` `r writer(data$Coordinate_Uncertainty)`).

>>>>

```{r load data}
data <- read.csv('SoS-ExampleCollection.csv') |>
  dplyr::mutate(
    Collection_number = as.numeric(Collection_number), 
    Coordinate_uncertainty = '+/- 5m') |>
  dplyr::filter(Collection_number == params$Collection_number)
```

@@

\vspace*{\fill}
\begingroup
`r data$Gen`. `r writer(data$Site)` `r data$latitude_dd` `r data$longitude_dd` (`r data$Datum` `r writer(data$Coordinate_uncertainty)`).

### 2025-09-20 LabelStyles/SoS-skeleton-large.Rmd
working on vignette example
reviewer comments on linux mac

```{r load data}
data <- read.csv('SoS-ExampleCollection.csv') |>
  dplyr::mutate(
    Collection_number = as.numeric(Collection_number), 
    Coordinate_Uncertainty = '+/- 5m') |>
  dplyr::filter(Collection_number == params$Collection_number)
```

\begin{center}
\textbf{\LARGE Seeds of Success}
\end{center}
\begin{flushright}
\uppercase{`r data$Family`}
\end{flushright}
\begingroup
\large
`r writer(paste(data$Genus, data$Epithet), italics = TRUE)` `r writer(data$Binomial_authority)` `r writer(data$Infrarank)` `r  writer(data$Infraspecies, italics = TRUE)` `r writer(data$Infraspecific_authority)`  
\normalsize
\endgroup

\vspace*{\fill}
\begingroup
`r data$Gen`. `r writer(data$Site)` `r data$latitude_dd` `r data$longitude_dd` (`r data$Datum` `r writer(data$Coordinate_Uncertainty)`).

>>>>

```{r load data}
data <- read.csv('SoS-ExampleCollection.csv') |>
  dplyr::mutate(
    Collection_number = as.numeric(Collection_number), 
    Coordinate_uncertainty = '+/- 5m') |>
  dplyr::filter(Collection_number == params$Collection_number)
```

\begin{center}
\textbf{\large Seeds of Success}
\end{center}
\begin{flushright}
\uppercase{`r data$Family`}
\end{flushright}
\begingroup
\large
`r writer(paste(data$Genus, data$Epithet), italics = TRUE)` `r writer(data$Binomial_authority)` `r writer(data$Infrarank)` `r writer(data$Infraspecies, italics = TRUE)` `r writer(data$Infraspecific_authority)`  
\normalsize
\endgroup

\vspace*{\fill}
\begingroup
`r data$Gen`. `r writer(data$Site)` `r data$latitude_dd` `r data$longitude_dd` (`r data$Datum` `r writer(data$Coordinate_uncertainty)`).

### 2025-09-20 LabelStyles/SoS-skeleton-noDir.Rmd
working on vignette example
reviewer comments on linux mac

\vspace*{\fill}
\begingroup
`r data$Gen`. `r writer(data$Site)` `r data$latitude_dd` `r data$longitude_dd` (`r data$Datum` `r writer(data$Coordinate_Uncertainty)`).

`r writer(data$Habitat)`. `r data$physical_environ`

Veg.: `r species_font(data$Vegetation)` `r associates_writer(data$Associates)`
`r writer(data$Notes)`  
\endgroup

>>>>

\vspace*{\fill}
\begingroup
`r data$Gen`. `r writer(data$Site)` `r data$latitude_dd` `r data$longitude_dd` (`r data$Datum` `r writer(data$Coordinate_uncertainty)`).

`r writer(data$Habitat)`. `r data$physical_environ`  
Veg.: `r species_font(data$Vegetation)` `r associates_writer(data$Associates)`
`r writer(data$Notes)`  
\endgroup

### 2025-09-20 manuscript/BarnebyLives An R package to create and qc herbarium labels.rmd
map example
reviewer comments on linux mac

Innovations in specimen digitization, data sharing, computing, DNA sequencing, and statistics have perhaps brought about greater use of herbarium specimens than ever before [@greve2016realising; @james2018herbarium; @brewer2019factors; @ronsted2020integrative]. 
The current use of specimens and their ancillary data extends well beyond their traditional roles in systematics and floristics, and studies utilizing collections are regularly carried out to better understand the ecological niches, phenological processes, and interactions of plants [@ronsted2020integrative; @davis2023herbarium]. 
We suspect that collections are yet to realize their full potential, and as currently novel approaches, such as electronic and remote sensing and meta-barcoding, become more accessible the use of collections will increase [@tosa2021rapid]. 
While image-based or purely observational (rather than collection-based) citizen science approaches (e.g., iNaturalist, BudBurst) have recently dovetailed with herbarium specimens to meet many current research needs, specimens contain rich data that are not accessible via images. 
Only specimens have the ability to: provide samples of DNA, secondary metabolites, or proteins, material for measuring (micro-)morphological attributes [@borges2020schrodinger], and seeds or pollen. 
These factors will ensure that the specimens remain the premier botanical data source into perpetuity.

@@

Currently, to our knowledge label generation functionality is provided explicitly by two programs, PLabel and Symbiota, and by the Microsoft Word tool Mail Merge [@perkins2020Plabel; @gries2014symbiota]. 
The office suite costs money, and in our experience, can be problematic to set up; further, its functionality ends with label creation. 
PLabel is a standalone program that has greatly enhanced functionality relative to a mail merge, allowing users to specify the layout and formatting of label components using an intuitive and local graphical user interface (GUI) functionality. 
However, beyond verifying the nations of collection it does not include data cleaning functionalities. 
While some sources indicate that it can only be used on Microsoft, we expect it to be usable on Linux and Mac using Windows 'emulators' like Wine. 
The increasingly popular Symbiota biodiversity data management software not only provides label generation capabilities but also provides data cleaning functionality in an attractive GUI web portal allowing for live management of collections and bypassing the need for a local installation, allowing it to be accessed on all operating systems. 
Symbiota offers functionality similar to the first four of our five stages of our 'Taxonomic' module and to our knowledge a check of the 'Political Boundaries’ (see Figure 1). 
However, not all herbaria use Symbiota and many have original database systems that they maintain (for example, Harvard University Herbarium, https://kiki.huh.harvard.edu/databases/specimen_index.html; Missouri Botanical Garden https://tropicos.org/specimen/Search; and The Consortium of Pacific Northwest Herbaria https://www.pnwherbaria.org/). 
However, and most importantly many collectors prefer to generate their own labels, especially as they are likely to send different sets of collections to different institutions. 
Accordingly, the functionality of Symbiota should exist in an ecosystem with alternative systems.  In scenarios where users want to keep rendering labels in either of the three existing alternatives, they can easily export data in the appropriate formats after utilizing BLs data cleaning utilities.

@@

# CONCLUSIONS
BarnebyLives is an R package that can be used to rapidly acquire relevant geographic and taxonomic data. 
It can also perform specialized spell checks and assorted curatorial tasks to produce both digital and analog data. 
The package relies on no licensed software, such as the Microsoft Office suite, and is suitable for install on all major operating systems  (Windows, Mac, Linux), however currently label generation support is only offered on Linux and Mac, with a small amount of use of the command line, which may be called from the Rstudio rather than a 'traditional' terminal.

>>>>

Innovations in specimen digitization, data sharing, computing, DNA sequencing, and statistics have perhaps brought about greater use of herbarium specimens than ever before [@greve2016realising; @james2018herbarium; @brewer2019factors; @ronsted2020integrative]. 
The current use of specimens and their ancillary data extends well beyond their traditional roles in systematics and floristics, and studies utilizing collections are regularly carried out to better understand the ecological niches, phenological processes, and interactions of plants [@ronsted2020integrative; @davis2023herbarium]. 
We suspect that collections are yet to realize their full potential, and as currently novel approaches, such as electronic and remote sensing and meta-barcoding, become more accessible the use of collections will increase [@tosa2021rapid]. 
While image-based or purely observational (rather than collection-based) citizen science approaches (e.g., iNaturalist, BudBurst) have recently dovetailed with herbarium specimens to meet many current research needs, specimens contain rich data that are not accessible via images. 
Of the two, specimens alone can provide samples of DNA, secondary metabolites, or proteins, material for measuring (micro-)morphological attributes [@borges2020schrodinger], and seeds or pollen. 
These factors will ensure that the specimens remain the premier botanical data source into perpetuity.

@@

Currently, to our knowledge label generation functionality is provided explicitly by two programs, PLabel and Symbiota, and by the Microsoft Word tool Mail Merge [@perkins2020Plabel; @gries2014symbiota]. 
The office suite costs money, and in our experience, can be difficult to set up; further, its functionality ends with label creation. 
PLabel is a standalone program that has greatly enhanced functionality relative to a mail merge, allowing users to specify the layout and formatting of label components using an intuitive and local graphical user interface (GUI) functionality. 
However, beyond verifying the nations of collection it does not include data cleaning functionalities. 
The increasingly popular Symbiota biodiversity data management software not only provides label generation capabilities but also provides data cleaning functionality in an attractive GUI web portal allowing for live management of collections and bypassing the need for a local installation, allowing it to be accessed on all operating systems. 
Symbiota offers functionality similar to the first four of our five stages of our 'Taxonomic' module and to our knowledge a check of the 'Political Boundaries’ (see Figure 1). 
However, not all herbaria use Symbiota and many have original database systems that they maintain (for example, Harvard University Herbarium, https://kiki.huh.harvard.edu/databases/specimen_index.html; Missouri Botanical Garden https://tropicos.org/specimen/Search; and The Consortium of Pacific Northwest Herbaria https://www.pnwherbaria.org/). 
However, and most importantly many collectors prefer to generate their own labels, especially as they are likely to send different sets of collections to different institutions. 
Accordingly, the functionality of Symbiota should exist in an ecosystem with alternative systems.  In scenarios where users want to keep rendering labels in either of the three existing alternatives, they can easily export data in the appropriate formats after utilizing BLs data cleaning utilities.

@@

# CONCLUSIONS
BarnebyLives is an R package that can be used to rapidly acquire relevant geographic and taxonomic data. 
It can also perform specialized spell checks and assorted curatorial tasks to produce both digital and analog data. 
The package relies on no licensed software, such as the Microsoft Office suite, and is suitable for install on all major operating systems  (Windows, Mac, Linux), however currently label generation support is only offered on Mac and Linux, with a small amount of use of the command line, which may be called from the Rstudio rather than a 'traditional' terminal.
